---
title: "Road accidents data analysis"
author: "Ahmed A. Radhi, Khalid Talal Suliman"
date: "3/13/2021"
output: html_document
---

```{r}
library(tidyverse)
```

# Task 1

```{r}
# Loading the data into R 
Accidents_2017 <- read_csv("Data/datagov-uk-6efe5505-941f-45bf-b576-4c1e09b579a1/2017-8.csv", 
col_names = c("Reference_Number","Easting","Northing","Vehicles_Num","Accident_Date","Time","Road_Class","Road_Surface",
              "Lightning_Cond","Weather_Cond","Vehicle_Type","Casualty_Class","Severity","Gender","Age"), skip = 1)

# Parsing the data
Accidents_Date <- Accidents_2017$Accident_Date
Accidents_Date <- parse_date(Accidents_Date, "%m/%d/%Y")

Time <- Accidents_2017$Time
Time <- parse_time(Time, "%H%M")

# Adding the new columns to the tibble.
Accidents_2017 <- Accidents_2017 %>%
  select(1:4, 7:15) %>%
  mutate(Accidents_Date, Time) 

Accidents_2017 <- Accidents_2017[,c(1:4,14,15,5:13)]
str(Accidents_2017)
```

We have 2,203 observations, 15 variables (with `Age` and `Road_Surface` as examples of continuous and categorical variables, respectively), and we have times and dates in the `Accidents_Date` and 'Time` variables.This satisfies the data requirements for the MATH 214 project.

# Description 
The dataset shows information on road accidents across Leeds during the year 2017.

 Reference_Number : Reference number of the accident
 Easting          : X-coordinate of the location of the accident                         using grid coordinates
 Northing         : Y-coordinate of the location of the accident                         using grid coordinates
 Vehicles_Num     : Number of vehicles involved in the accident
 Accident_Date    : Date of the accident 
 Time             : Time of the accident
 Road_Class       : Class of road on which the accident happened
 Road_Surface     : Surface of the road during the accident
 Lightning_Cond   : Light conditions during the accident
 Weather_Cond     : Weather conditions during the accident 
 Vehicle_Type     : Type of vehicle that caused the accident
 Casualty_Class   : Class of the casualty
 Severity         : Degree of severity of the accident
 Gender           : Gender of the driver
 Age              : Age of the driver 

# Reference 
https://data.world/datagov-uk/6efe5505-941f-45bf-b576-4c1e09b579a1

# Task 2

