---
title: "Road accidents data analysis"
author: "Ahmed A. Radhi, Khalid Talal Suliman"
date: "3/13/2021"
output: html_document
---

```{r}

packages <- c("tidyverse", "leaflet","sf")
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

library(tidyverse)
library(leaflet)
library(sf)
```

# Task 1

```{r}
# Loading the data into R 
Accidents_2017 <- read_csv("Data/datagov-uk-6efe5505-941f-45bf-b576-4c1e09b579a1/2017-8.csv", 
col_names = c("Reference_Number","Easting","Northing","Vehicles_Num","Accident_Date","Time","Road_Class","Road_Surface",
              "Lightning_Cond","Weather_Cond","Vehicle_Type","Casualty_Class","Severity","Gender","Age"), skip = 1)

# Parsing the data
Accidents_Date <- Accidents_2017$Accident_Date
Accidents_Date <- parse_date(Accidents_Date, "%m/%d/%Y")

Time <- Accidents_2017$Time
Time <- parse_time(Time, "%H%M")

# Adding the new columns to the tibble.
Accidents_2017 <- Accidents_2017 %>%
  select(1:4, 7:15) %>%
  mutate(Accidents_Date, Time) 

Accidents_2017 <- Accidents_2017[,c(1:4,14,15,5:13)]

```

```{r}
str(Accidents_2017)
```


We have 2,203 observations, 15 variables (with `Age` and `Road_Surface` as examples of continuous and categorical variables, respectively), and we have times and dates in the `Accidents_Date` and `Time` variables. This satisfies the data requirements for the MATH 214 project.

## Description 
The dataset shows information on road accidents across Leeds during the year 2017.

 Reference_Number : Reference number of the accident (character)
 
 Easting          : X-coordinate of the location of the accident                         using grid coordinates (double)
 
 Northing         : Y-coordinate of the location of the accident                         using grid coordinates (double)
 
 Vehicles_Num     : Number of vehicles involved in the accident                          (double)
 
 Accident_Date    : Date of the accident (character)
 
 Time             : Time of the accident (double)
 
 Road_Class       : Class of road on which the accident happened                         (character)
 
 Road_Surface     : Surface of the road during the accident                              (character)
 
 Lightning_Cond   : Light conditions during the accident
                     (character)
                     
 Weather_Cond     : Weather conditions during the accident 
                     (character)
                     
 Vehicle_Type     : Type of vehicle that caused the accident
                     (character)
                     
 Casualty_Class   : Class of the casualty (character)
 
 Severity         : Degree of severity of the accident (character)
 
 Gender           : Gender of the driver (character)
 
 Age              : Age of the driver (double)

## Reference 
https://data.world/datagov-uk/6efe5505-941f-45bf-b576-4c1e09b579a1

# Task 2

## Variance

```{r}
# Variance (Age)

ggplot(data = Accidents_2017, mapping= aes(x=Age)) +
  geom_histogram(binwidth = 1,fill="black", color = "white") +
  ggtitle("Number of Casualties Based on Age") + 
  ylab("Number of Casualties")

Accidents_2017 %>%
  group_by(Age) %>%
  summarise(count = n())

```

As the age increase,the number of casualties increases until the age 26. After the age 26 , as the age increase the number of casualties decreases. This was expected as older drivers are more experienced than younger ones. It is a important to note that there is a big jump in the number of casualties from age 17 to 18.


```{r}
# Variance (Gender)

ggplot(data = Accidents_2017, mapping = aes(x=Gender)) + 
  geom_bar(aes(fill = Gender) , width = 0.5) +
  ggtitle("Number of Casualties Based on Gender") + 
  ylab("Number of Casualties")

Accidents_2017 %>%
  group_by(Gender) %>%
  summarise(count = n())
```

Males have a higher tendency to cause accidents.The ratio between males and females in accidents is slightly above 1.3.

```{r}
# Variance (Number of Accidents Per Month)

# Accidents per day

Accidents_2017 %>%
  group_by(Accidents_Date) %>%
  summarise(sum = n())

# Accidents per month

( Accidents_by_month <- 
    Accidents_2017 %>%
    mutate(month = format(Accidents_Date,"%m")) %>%
    group_by(month) %>%
    summarise(accidents_num = n())
)

# Plot of Accidents per month
ggplot(data = Accidents_by_month, mapping = aes(x = month, y = accidents_num)) +
  geom_bar(aes(fill = month) , show.legend = F, stat = "identity") +
  ggtitle("Number of Accidents Per Month") +
  xlab("Month") +
  ylab("Number of Accidents")

```

Throughout the year, there were at least 150 accidents every month. February was the month with the least number of accidents, a reason for this is February 2017 had only 28 days. On the other hand, November had the most accidents (234). Towards the end of the year in Leeds the number of accidents increase.


```{r}
# Variance (Road Surface)
ggplot(data = Accidents_2017) +
  geom_bar(mapping = aes(x = Road_Surface,fill = Road_Surface))

Accidents_2017 %>%
  group_by(Road_Surface) %>%
  summarise(count = n())
```

Most accidents happen when the road is dry which is expected as drivers take more care when the road is bad. On average, Leeds has 175.4 rainfall days and 10.6 snowfalls days per year so for the remaining days the road is dry. Thus number of accidents per day for each climate is 2.97 when it is raining, 3.3 when it is snowing and 9.2 when the climate is dry. The number of accidents per day when the weather is dry is extremely highly and shows how careless Leeds' drivers are the road surface is dry.



```{r}
# Variance (Time)
ggplot(Accidents_2017,mapping=aes(x=Time))+
  geom_histogram(binwidth = 2000)+
  ggtitle("Accidents against Time")+
  ylab("Number of Accidents")+
  xlab("Time")

```

From 12 am till 6 am the number of accidents is remarkably low. After 6 am the number of accidents keeps increasing till 5.30 pm. After that the number of accidents keep decreasing. It is noteworthy to mention that there are 2 big drops in the number of accidents which  could represent something. 

## Covariance

```{r}

# Covariance (Age/Gender against Count)

Accidents_2017 %>%
  group_by(Gender, Age) %>%
  summarise(Count = n() ) %>%
  ungroup() %>%
  ggplot(mapping = aes(x=Age , y = Count)) +
  geom_boxplot(aes(color = Gender))

Accidents_2017 %>%
  group_by(Gender, Age) %>%
  summarise(Count = n() )

```

```{r}
# Covariance (Time and Severity)
ggplot(data = Accidents_2017)+ 
  geom_boxplot(mapping = aes(x = reorder(Severity, Time, FUN = median), y = Time))+ coord_flip()+
  ggtitle("Severity against Time")+
  ylab("Severity")+
  xlab("Time")
```

Most serious and slight accidents happen between 10 am and 17.30 pm. While most fatal accidents occur between 16 pm and 22 pm. Serious accidents have the widest inter-quartile range while fatal accidents have the smallest inter=quartile range.   

```{r}
# Covariance(Age and Severity)
ggplot(data = Accidents_2017)+ 
  geom_boxplot(mapping = aes(x = reorder(Severity, Age, FUN = median), y = Age))+ coord_flip()+
  ggtitle("Severity against Age")+
  ylab("Age")+
  xlab("Severity")
```

Fatal accidents tend to happen to people under the age of 25 more frequently.Slight accidents have a few outliers. Serious accidents have the widest inter-quartile range.

```{r}
Accidents_2017 %>%
  group_by(Severity) %>%
  summarise(count = n())

```

The number of accidents that happen between 12 am and 7 am are low.

Generally, after 7 am as the time increases the number of accidents increase till 6 pm. After 6 pm the number of accidents decrease. There are 2 low drops which might represent something. 

```{r}
Accidents_2017 %>%
  group_by(Weather_Cond) %>%
  summarise(count = n())
```

## Extra Plot: 

### Locations of the Accidents

```{r}

long_lat <- Accidents_2017 %>%
  st_as_sf(coords = c("Easting", "Northing"), crs = 27700) %>%
  st_transform(4326) %>%
  st_coordinates() %>%
  as_tibble()

long_lat_FULL <- Accidents_2017 %>%
  mutate(long_lat)  %>%
  select(Reference_Number, X, Y, Severity)
  
pal <- colorFactor(c("yellow", "orange", "red"), domain = c("Slight", "Serious", "Fatal"))

leaflet(data = long_lat_FULL)%>%
  addTiles() %>%
  addCircleMarkers(lng = ~X, lat = ~Y, 
                   color = ~pal(Severity),
                   stroke = F, fillOpacity = 0.75,
                   clusterOptions = markerClusterOptions())  %>%
  addLegend(colors = c("yellow", "orange", "red") , labels = c("Slight", "Serious", "Fatal"), title = "Severity")

             
```



# Questions

1. What are the worst conditions to drive in ?
2. What season has the most accidents ?
3. What is the worst time to drive at ?
4. Which area in Leeds has the most accidents ?
5. Which roads are the most dangerous ?


